---
title: "Lab 02: A Data Scientist's MOMA Tour"
subtitle: "CS631"
author: "Alison Hill"
output:
  html_document:
    highlight: pygments
    theme: flatly
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    number_sections: TRUE
    code_folding: hide
---
```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, comment = NA, warnings = FALSE, errors = FALSE, messages = FALSE, tidy = FALSE, cache = FALSE)
```

# Goals for Lab 02

# Inspiration + data

https://fivethirtyeight.com/features/a-nerds-guide-to-the-2229-paintings-at-moma/

https://github.com/MuseumofModernArt/collection

https://medium.com/@foe/here-s-a-roundup-of-how-people-have-used-our-data-so-far-80862e4ce220

# Read in the data

Note! This is not the original data- I did a lot of cleaning and decision-making in the pre-processing. The below contains only paintings and drawings in the MOMA collection.

```{r}
library(here)
library(readr)
library(janitor)
library(dplyr)
moma <- read_csv(here::here("data", "artworks-cleaned.csv"))
```

# Explore with `dplyr`

## How many paintings?

```{r}
library(dplyr)
moma
glimpse(moma)
```

## How many paintings by year acquired?
```{r}
moma %>% 
  count(year_acquired, sort = TRUE)
```

## How many paintings by year created?

```{r}
moma %>% 
  count(year_created, sort = TRUE)
```

## How many paintings since 2000?

```{r}
moma %>% 
  filter(year_acquired >= 2000) %>% 
  tally()
```


## Most popular artists

```{r}
moma %>% 
  count(artist, sort = TRUE)
```


## What about dimensions?

```{r}
library(ggplot2)
ggplot(moma, aes(x = height_cm, y = width_cm)) +
  geom_point()
```

Let's try filtering

```{r}
moma_for_size <- moma %>% 
  filter(height_cm < 600, width_cm < 760)
ggplot(moma_for_size, aes(x = height_cm, y = width_cm)) +
  geom_point(alpha = .1)
```


# Year painted vs year acquired

```{r}
ggplot(moma, aes(year_created, year_acquired)) +
  geom_point(alpha = .1) +
  geom_abline(intercept = 0, colour = "red") +
  coord_fixed()
```



## How many feature males only?

```{r}
moma_solo <- moma %>% 
  filter(num_artists == 1)
ggplot(moma_solo, aes(year_created, year_acquired)) +
  geom_point(alpha = .1) +
  geom_abline(intercept = 0, colour = "red") +
  coord_fixed() +
  facet_wrap(~artist_gender)
```



## How many feature females only?
