---
title: "Lab 05: Addendum"
subtitle: "CS631"
author: "Alison Hill"
output:
  html_document:
    theme: flatly
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    number_sections: TRUE
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, comment = NA, warning = FALSE, errors = FALSE, message = FALSE, tidy = FALSE, cache = FALSE)
```

# Packages

```{r}
library(tidyverse)
library(babynames)
# load hd data
hot_dogs_aff <- read_csv("http://bit.ly/cs631-hotdog-affiliated", 
    col_types = cols(
      affiliated = col_factor(levels = NULL), 
      gender = col_factor(levels = NULL)
      )) %>% 
  mutate(post_ifoce = year >= 1997) %>% 
  filter(year >= 1981 & gender == "male") 
```

# Plot 1: Add

What is this code trying to accomplish?

```{r}
temp=hot_dogs_aff

for (i in 1:nrow(hot_dogs_aff)) { 
  # if the maximum num_eaten is equal to the num_eaten for the year AND it's not the same as the year before
  if ((max(temp$num_eaten) == hot_dogs_aff$num_eaten[i]) && (max(temp$num_eaten) != hot_dogs_aff$num_eaten[i+1]) && (i<nrow(hot_dogs_aff))) {
    hot_dogs_aff$record[i] = TRUE
    temp = temp[-1,] # take the most recent year OUT of the running 
 } else if (i<nrow(hot_dogs_aff)) {
   hot_dogs_aff$record[i] = FALSE
   temp = temp[-1,]
 }
}

label.df <- data.frame(year = hot_dogs_aff$year[hot_dogs_aff$record],
                       num_eaten = hot_dogs_aff$num_eaten[hot_dogs_aff$record]) # turn this into a label dataframe
```

# Tidyverse approach

The first thing we notice is that we don't have data about whether each year's winner is a record or not. 

```{r}
hot_dogs <- read_csv("http://bit.ly/cs631-hotdog", 
    col_types = cols(
      gender = col_factor(levels = NULL)
    ))
```


Since our data is nicely tidy, we can use `dplyr` window functions:

- First, we use base R's `cummax` to create a new variable that reflects the maximum HDB eaten cumulatively, that is, compared to all earlier years. For this reason, the `arrange(year)` here is critical.

- Next, we want to know if the `hdb_record` is actually a *new* record or not, compared to all previous years. We can use `case_when` to create a logical variable that is TRUE if the `hdb_record` for a given year is greater than the `hdb_record` from the year before (using `dplyr::lag`). If not, this variable is FALSE.

```{r}
hot_dogs_records <- hot_dogs %>% 
  filter(year >= 1980 & gender == 'male') %>%  
  arrange(year) %>% 
  mutate(hdb_record = cummax(num_eaten),
         new_record = case_when(
           hdb_record > lag(hdb_record) ~ TRUE,
           TRUE ~ FALSE
         )) %>% 
  filter(year >= 1981)
```


We'll also make our x-axis ticks again...


```{r}
years_to_label <- seq(from = 1981, to = 2017, by = 4)
years_to_label
hd_years <- hot_dogs_records %>%
  distinct(year) %>% 
  mutate(year_lab = ifelse(year %in% years_to_label, year, ""))
```


```{r}
hdb_records <- ggplot(hot_dogs_records, 
                      aes(x = year, y = num_eaten)) + 
  geom_col(aes(fill = new_record)) + 
  labs(x = "Year", y = "Hot Dogs and Buns Consumed") +
  ggtitle("Nathan's Hot Dog Eating Contest Results, 1981-2017") +
  scale_fill_manual(values = c('#284a29', '#629d62')) + 
  scale_y_continuous(expand = c(0, 0),
                     breaks = seq(0, 70, 10)) +
  scale_x_continuous(expand = c(0, 0), 
                     breaks = hd_years$year,
                     labels = hd_years$year_lab) + 
  coord_cartesian(xlim = c(1980, 2018), ylim = c(0, 80)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text = element_text(size = 12),
        panel.background = element_blank(),
        axis.line.x = element_line(color = "gray92", 
                                   size = 0.5),
        axis.ticks = element_line(color = "gray92", 
                                  size = 0.5),
        text = element_text(family = "Lato"),
        legend.position = "bottom",
        panel.grid.minor = element_blank())
hdb_records
```

# Plot 2

Goal is to observe popular first letter trends in `babynames`.
First filter/arrange data.
```{r}
# add column of first letters
baby_letters <- babynames %>% 
  mutate(first_letter = str_sub(name, 1, 1))

# add up proportion for each letter by year + sex
letter_by_year <- baby_letters %>%
  count(year, sex, first_letter, wt = prop) %>% 
  rename(total_prop = nn)

# label most popular for each year + sex
letter_by_year <- letter_by_year %>% 
  group_by(sex, year) %>% 
  mutate(top_yearly = case_when(
    total_prop == max(total_prop) ~ first_letter)) %>% 
  ungroup() %>% 
  mutate(top_latest = case_when(
      year == max(year) ~ top_yearly
    ))

# make a data frame with just latest top letters
top_current <- letter_by_year %>% 
  filter(!is.na(top_latest)) %>% 
  rename(color_by = top_latest) %>% 
  select(sex, first_letter, color_by)

# now merge in
letter_by_year <- letter_by_year %>% 
  left_join(top_current) %>% 
  mutate(color_by = replace_na(color_by, "else"))

# set up my colors for males and females
my_colors <- c("#FF1493", "gray80", "#3399ff")
```

Plot it!

```{r}
ggplot(letter_by_year, aes(x = year, y = total_prop, color = color_by, group = first_letter)) +
  geom_line() +
  scale_color_manual(values = my_colors) +
  facet_wrap(~sex)
```

```{r}
ColorM2 <- c(A="Gray80",B="gray80",C="gray80",D="gray80",E="gray80",F="gray80",G="gray80",H="gray80",I="gray80",J="Black",K="gray80",L="gray80",M="gray80",N="gray80",O="gray80",P="gray80",Q="gray80",R="gray80",S="gray80",T="gray80",U="gray80",V="gray80",W="gray80",X="gray80",Y="gray80",Z="gray80")

ColorF2 <- c(A="Black",B="gray80",C="gray80",D="gray80",E="gray80",F="gray80",G="gray80",H="gray80",I="gray80",J="gray80",K="gray80",L="gray80",M="gray80",N="gray80",O="gray80",P="gray80",Q="gray80",R="gray80",S="gray80",T="gray80",U="gray80",V="gray80",W="gray80",X="gray80",Y="gray80",Z="gray80")
```

