---
title: "In-Class Slides & Reading for CS631 Labs"
output: html_document
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, comment = NA, warnings = FALSE, errors = FALSE, messages = FALSE, tidy = FALSE, echo = FALSE)
```

The following are all *free* books, available to read online:

- Modern Dive: http://moderndive.com
- R for Data Science: http://r4ds.had.co.nz
- Fundamentals of Data Visualization: http://serialmentor.com/dataviz/
- Data Visualization: A Practical Introduction: http://socviz.co

## Slide Decks & Reading {.tabset}

### 01: Working in R

```{r}
knitr::include_url("slides/01-slides.html")
```

- R for Data Science, Workflow: Basics Chapter: http://r4ds.had.co.nz/workflow-basics.html

- Modern Dive, Getting Started Chapter: http://moderndive.com/2-getting-started.html

- R & RStudio Basics: https://bookdown.org/chesterismay/rbasics/3-rstudiobasics.html

- RStudio IDE Cheatsheet: https://github.com/rstudio/cheatsheets/blob/master/rstudio-ide.pdf

### 02: Working in the Tidyverse

```{r}
knitr::include_url("slides/02-slides.html")
```

- R for Data Science, Data Visualization Chapter: http://r4ds.had.co.nz/data-visualisation.html

- Modern Dive, Data Visualization Chapter: http://moderndive.com/3-viz.html

- RStudio `ggplot2` Cheatsheet: https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf

### 03: Working with Data

```{r}
knitr::include_url("slides/03-slides.html")
```

### 04: Working with Tidy Data

```{r}
knitr::include_url("slides/04-slides.html")
```

### 06: Slides

```{r}
knitr::include_url("slides/06-slides_xaringan.html")
```

### 06+07: Blogdown

```{r}
knitr::include_url("slides/06-slides_blogdown.html")
```

```{r}
knitr::include_url("slides/07-prework-slides_blogdown.html")
```

```{r}
knitr::include_url("slides/07-slides_blogdown.html")
```

### 11: 11 ways

```{r}
knitr::include_url("slides/11-slides.html")
```

### 12: Wrap-up

```{r}
knitr::include_url("slides/12-slides.html")
```

