---
title: "Lab 01: CS631"
subtitle: "R Basics"
author: "Alison Hill"
output:
  xaringan::moon_reader:
    css: ["default", "css/ohsu.css", "css/ohsu-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
  comment = NA, dpi = 300,
  fig.align = "center", out.width = "80%", cache = FALSE)
library(tidyverse)
library(babynames)
```

# R Basics

--

* R is an interpreter `>`

--

* You are an object maker & borrower `<-`

--

* Know your data types

--

* Case matters (`i_like_snake_case`)

--

* Use comments! All text behind a `#` is a comment

--

* Use packages; think: *"install once per machine, load once per R session"*

--

* Use the `%>%`; think: *"dataframe first, dataframe once"*

---

## R is an interpreter `>`

--

You enter commands line-by-line (as opposed to compiled languages).

--

* The `>` means R is a ready for a command

--

* The `+` means your last command isn't complete

--

    - If you get stuck with a `+` use your escape key!

---

## You are an object maker


```{r}
data_vis <- c("Steven", "Jackie", "Alison") # c = combine
data_vis
num_labs <- c(1:10)
num_labs
mood <- rep("yippee", sum(num_labs)) # replicate X times
mood
```

---

## You are an object borrower


```{r}
letters # built-in
my_names <- babynames # from the babynames package
my_names
```



---

## Every object has a class

Classes:

* Numeric
* Character
* Factor
* Logical
* Double (`?double`: "identical to numeric", but "double precision")
* a few others you may run into `?typeof`

```{r}
typeof(num_labs) # numeric
typeof(mood) # "yippee" is a character
```

Vectors can only contain one type of thing.

```{r mixing-types}
my_things <- c("oh", 1, 2, "hi") 
my_things
typeof(my_things) 
```

R denotes missing data with a special type of thing, `NA`, which is not a character (and hence not in quotes). It is actually a logical, which returns TRUE or FALSE.

```{r missing-values-in-vector}
?logical
missing <- c(NA, NA, NA)
typeof(missing)
is.na(missing)
dollars <- c(12, 1, 2, 3, NA)
typeof(dollars)
is.na(dollars)
dollars*3
mean(dollars)
mean(dollars, na.rm = TRUE)
```


---

## Case matters

```{r mean-works}
# calculate the mean
mean(dollars) # calculate mean of numbers 
```

```{r mean-wont-work, eval = FALSE}
Mean(dollars) # no function
mean(Dollars) # no data 
```

Read more: http://r4ds.had.co.nz/workflow-basics.html#whats-in-a-name

---

## Text behind a `#` is a comment

```{r}
2+2 # add 2 + 2 here

# I can say anything I want here
```


---

## Packages! 

**Install once** per machine

```{r eval = FALSE}
install.packages("dplyr")
```

**Load once** per R work session

```{r eval = FALSE}
library(dplyr)
```

*also: quotes matter, sorry*

This can get to be a long list if we want to use a lot of packages in our work session. We can make a function to load a *list* of packages, and install them if not already installed (more on functions later).

```{r package-list-to-load}
pkgs <- c("readr", "readxl", "haven", "dplyr", "tidyr", "stringr") # list packages needed
```

```{r}
ipak <- function(pkg){
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg)) 
        install.packages(new.pkg, dependencies = TRUE)
    sapply(pkg, require, character.only = TRUE)
}
```

```{r run-pkg-function, eval = FALSE}
ipak(pkgs) # take function, and give it that list
```

---

## Functions 

Sometimes abbreviated `funs` in documentation, which is a little ironic 😉.

Some simple ones:

```{r}

class_weeks <- seq(1, 12, 1)
class_labs <- seq(1, 12, 2)
```


Read more: http://r4ds.had.co.nz/workflow-basics.html#calling-functions

---

## Piping/chaining with the `%>%` operator

This is the magrittr operator: `%>%`. You can use with the `dplyr` package loaded (which we just did above).

You can check via `sessionInfo()`
```{r}
sessionInfo()
```

---

## *Nesting* your dataframe in commands is hard to read

```{r}
head(iris)
```

## *Piping* your dataframe into a command lets you read L to R

```{r}
iris %>% head(.)
```

---

## Sequences of commands are read *inside out*

```{r}
head(iris[iris$Species == "virginica", ])
```

### Chaining your commands lets you read *L to R*
```{r}
iris %>% filter(Species == "virginica") %>% head(.) 
```

The `(.)` is implicit- try the above code without the parentheses or the `.`. But keep in mind that under the hood, this is what the implied syntax is. It will be useful later on!

