---
title: "Lab 02: CS631"
subtitle: "Tidyverse Basics"
author: "Alison Hill"
output:
  xaringan::moon_reader:
    css: ["default", "css/ohsu.css", "css/ohsu-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
  comment = NA, dpi = 300, echo = FALSE,
  fig.align = "center", out.width = "80%", cache = TRUE)
library(tidyverse)
```

# Tidyverse basics

In David Robinson's DataCamp course you learned:

- `<-` (variable assignment)
- `%>%` (then...)
- `dplyr`, `ggplot2` (packages)
  - `install.packages("dplyr")` (1x per machine)
  - `library(dplyr)` (1x per work session)

---
class: center, middle, inverse

## Let's review

---
# Data for today

We'll use data from the Museum of Modern Art (MOMA)

- Publicly available on [GitHub](https://github.com/MuseumofModernArt/collection)
- As analyzed by [fivethirtyeight.com](https://fivethirtyeight.com/features/a-nerds-guide-to-the-2229-paintings-at-moma/)
- And by [others](https://medium.com/@foe/here-s-a-roundup-of-how-people-have-used-our-data-so-far-80862e4ce220)

---
# Get the data

Use this code chunk to import my cleaned CSV file:

```{r}
library(here)
library(readr)
moma <- read_csv("http://bit.ly/cs631-moma")
```


---

# From DataCamp Chapter 1

all `dplyr`

- print a tibble
- `filter`
- `arrange`
- `mutate`

---

# From Lab 01

all `dplyr`

- `glimpse`
- `distinct`
- `count`


---
# Lab 02: Challenge 1

- How many rows/observations are in `moma`?
- How many variables are in `moma`?
- What is the earliest year
- How many distinct artists are there?
- Which artist has the most paintings ever owned by `moma`? How many?


---
# From Lab 01: Challenge 1

```{r}
moma
glimpse(moma)
```


---

# From DataCamp Chapter 2

all `ggplot2`

- `aes(x = , y = )` (aesthetics)
- `aes(x = , y = , color = )` (add color)
- `aes(x = , y = , color = , size = )` (add size)
- `+ facet_wrap(~ )` (facetting)
