---
title: "Lab 02: CS631"
subtitle: "Tidyverse Basics"
author: "Alison Hill"
output:
  xaringan::moon_reader:
    css: ["default", "css/ohsu.css", "css/ohsu-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
  comment = NA, dpi = 300,
  fig.align = "center", out.width = "80%", cache = FALSE)
library(tidyverse)
```

# Tidyverse basics

In David Robinson's DataCamp course you learned:

- `<-` (variable assignment)
- `%>%` (then...)
- `dplyr`, `ggplot2` (packages)
  - `install.packages("dplyr")` (1x per machine)
  - `library(dplyr)` (1x per work session)

---
class: center, middle, inverse

## Let's review

---
# Data for today

We'll use data from the Museum of Modern Art (MoMA)

- Publicly available on [GitHub](https://github.com/MuseumofModernArt/collection)
- As analyzed by [fivethirtyeight.com](https://fivethirtyeight.com/features/a-nerds-guide-to-the-2229-paintings-at-moma/)
- And by [others](https://medium.com/@foe/here-s-a-roundup-of-how-people-have-used-our-data-so-far-80862e4ce220)

---
# Get the data

Use this code chunk to import my cleaned CSV file:

```{r}
library(readr)
moma <- read_csv("http://bit.ly/cs631-moma")
```


---

# Data wrangling so far

All functions from `dplyr` package

.pull-left[
From DataCamp Chapter 1

- print a tibble

- `filter`

- `arrange`

- `mutate`
]

--

.pull-right[
From Lab 01


- `glimpse`

- `distinct`

- `count`
]

---
# Lab 02: Challenge 1

Use all the `dplyr` functions you have learned to answer the following questions:

1. How many rows/observations are in `moma`?
1. How many variables are in `moma`?
1. What is the first painting acquired by MoMA? Which year? Which artist? What title?
    - *Hint: you may want to look into `select` + `arrange`*
1. What is the oldest painting in the collection? Which year? Which artist? What title? *(see above hint)*
1. How many distinct artists are there?
1. Which artist has the most paintings ever owned by MoMA? How many paintings are by this artist?

If you want more:
1. In what year were the most paintings acquired? Created?
1. In what year was the first painting by a (solo) female artist acquired? When was that painting created? Which artist? What title? 



---

# From DataCamp Chapter 2

all `ggplot2`

- `aes(x = , y = )` (aesthetics)
- `aes(x = , y = , color = )` (add color)
- `aes(x = , y = , size = )` (add size)
- `+ facet_wrap(~ )` (facetting)

---
# Old School (Challenge 2)<sup>1</sup>

- Sketch the graphics below on paper, where the `x`-axis is variable `year_created` and the `y`-axis is variable `year_acquired`

```{r echo=FALSE}
simple_ex <-
  data_frame(
    year_created = c(1980, 1990, 2000, 2010),
    year_acquired = c(1985, 1995, 2005, 2015),
    area_cmsq = c(3, 2, 1, 2),
    artist_gender = c("male", "male", "female", "female")
  )
simple_ex
```

<!-- Copy to chalkboard/whiteboard -->

1. A scatter plot
1. A scatter plot where the `color` of the points corresponds to `artist_gender`
1. A scatter plot where the `size` of the points corresponds to `area_cmsq`

.footnote[
[1] Shamelessly borrowed with much appreciation to [Chester Ismay](https://ismayc.github.io/talks/ness-infer/slide_deck.html)
]

---

# 1. A scatterplot

```{r, eval=FALSE}
library(ggplot2)
ggplot(data = simple_ex, mapping = aes(x = year_created, 
                                       y = year_acquired)) + 
  geom_point()
```
--

```{r, echo=FALSE, fig.height=4.4}
ggplot(data = simple_ex, aes(x = year_created, 
                             y = year_acquired)) + 
  geom_point()
```


---


# 2. `color` points by `D`

```{r, eval=FALSE}
library(ggplot2)
ggplot(data = simple_ex, mapping = aes(x = year_created, 
                                       y = year_acquired)) + 
  geom_point(mapping = aes(color = artist_gender))
```
--

```{r, echo=FALSE, fig.height=4.4}
ggplot(data = simple_ex, mapping = aes(x = year_created, 
                                       y = year_acquired)) + 
  geom_point(mapping = aes(color = artist_gender))
```


---

# 3. `size` points by `C`

```{r, eval=FALSE}
library(ggplot2)
ggplot(data = simple_ex, mapping = aes(x = year_created, 
                                       y = year_acquired, 
                                       size = artist_gender)) + 
  geom_point()
```
--

```{r, echo=FALSE, fig.height=4.4}
ggplot(data = simple_ex, mapping = aes(x = year_created, 
                                       y = year_acquired, 
                                       size = area_cmsq)) + 
  geom_point()
```

---

# [The Five-Named Graphs](http://moderndive.com/3-viz.html#FiveNG)


- Scatterplot: `geom_point()`
- Line graph: `geom_line()`
- Histogram: `geom_histogram()`
- Boxplot: `geom_boxplot()`
- Bar graph: `geom_bar()`



---
# Lab 02: Plotting Challenges

Challenges 3 & 4 are in the [lab](https://apreshill.github.io/data-vis-labs-2018/02-moma.html)!

https://apreshill.github.io/data-vis-labs-2018/02-moma.html
