---
title: "Lab 04: CS631"
subtitle: "Working with Tidy Data"
author: "Alison Hill"
output:
  xaringan::moon_reader:
    css: ["default", "css/ohsu.css", "css/ohsu-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
  comment = NA, dpi = 300,
  fig.align = "center", out.width = "80%", cache = FALSE)
library(tidyverse)
library(here)
```


class: middle, center, inverse

# ⌛️

## Let's review 

---

## Data wrangling to date! 


.pull-left[

From `dplyr`:
- `filter`
- `arrange`
- `mutate`
- `group_by`
- `summarize`
- `glimpse`
- `distinct`
- `count`
- `tally`
- `pull`
- `top_n`

]

--

.pull-right[
Let's add from `dplyr`:

- `select`
- `rename`
- `recode`
- `case_when`

From `tidyr`:

- `gather`
- `separate`
- `spread`
- `unite`

Plus 1 other package:

- `skimr::skim`

]

---

# Un-tidy cakes

```{r echo = FALSE}
cakes_untidy <- read_csv(here::here("data", "tidy-gbbo-cakes.csv"), col_types = cols(series = col_factor(levels = NULL))) %>% 
  count(series, challenge, cake) %>% 
  drop_na(cake) %>% 
  rename(bakes = n) %>% 
  spread(cake, bakes, fill = 0) 

cakes_split <- cakes_untidy %>% 
  split(.$series) 
```


.pull-left[
```{r echo = FALSE}
cakes_split[[1]]
cakes_split[[2]]
```
]

--


.pull-right[
```{r echo = FALSE}
cakes_split[[3]]
cakes_split[[4]]
```
]

---

# Still un-tidy cakes


```{r}
cakes_untidy %>% 
  bind_rows()
```


---

# Finally tidy cakes

```{r}
cakes_tidy <- cakes_untidy %>% 
  gather(bake_type, num_bakes, cake:pie_tart, 
         factor_key = TRUE) %>% 
  arrange(series)
cakes_tidy
```

---
class: middle, inverse, center

## Know Your Tidy Data

---

```{r}
glimpse(cakes_tidy)
```


---

```{r}
library(skimr)
skim(cakes_tidy)
```

---

```{r}
skim(cakes_tidy) %>% 
  summary() 
```


---
class: middle, inverse, center

## Benefits of Tidy Data


---


```{r}
cakes_tidy %>% 
  count(challenge, bake_type, wt = num_bakes, sort = TRUE)
```

---


```{r}
cakes_tidy %>% 
  count(series, bake_type, wt = num_bakes)
```

---

```{r}
library(skimr)
cakes_tidy %>% 
  group_by(bake_type) %>% 
  select_if(is.numeric) %>% 
  skim()
```

See: https://suzanbaert.netlify.com/2018/01/dplyr-tutorial-1/

---


```{r}
cakes_by_series <- cakes_tidy %>% 
  count(series, bake_type, wt = num_bakes) 
cakes_by_series
```

---


```{r fig.height = 3, fig.width = 4}
ggplot(cakes_by_series, aes(x = series, y = n, 
                            color = bake_type,
                            group = bake_type)) +
  geom_point() +
  geom_line() +
  expand_limits(y = 0)
```

---
class: middle, inverse, center


## You have 2 challenges today!

Described [here](04-challenge.html)
Reference lab [here](04-distributions.html)


---
class: middle, inverse, center

# 🍱

## Tidy Data:

http://r4ds.had.co.nz/tidy-data.html

http://moderndive.com/4-tidy.html

http://vita.had.co.nz/papers/tidy-data.html

https://github.com/jennybc/lotr-tidy#readme